### 2023-09-06 doc/_bayesian_inference.qmd
making further progress on several sections

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $T \times B$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique mutant barcodes $M$ and the
number of unique neutral barcodes $N$, and $T$ is the number of time points
where measurements were taken. The data matrix is then of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
$${#eq-R_mat}
where each row $\underline{r}_t$ is a $B$-dimensional array containing the
raw barcode counts at time $t$. We can further split each vector
$\underline{r}_t$ into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \;
\underline{r}_t^{M}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the neutral lineages barcode counts
$\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
counts $\underline{r}_t^{M}$. Our objective is to compute the joint probability
distribution of all relative fitness values given our data. We can express this 
joint distribution using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function unless
otherwise stated. We can ignore the denominator on the right-hand side of
@eq-bayes_obj, since it serves as a normalization constant that does not depend
on our parameter of interest, simply writing
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)
$${#eq-bayes_obj_propto}

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. We can always marginalize the joint
distribution to obtain the distribution we are interested in. To include these
parameters, let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R_mat, each row of $\underline{\underline{F}}$
can be split into two components of the form
$$
\underline{f}_t = \begin{bmatrix}
\underline{f}_t^M \;
\underline{f}_t^N
\end{bmatrix},
$${#eq-f_vec}
to separate the corresponding mutant and neutral barcode frequencies. With these
variables in hand, the full inference problem we must solve takes the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
We can numerically integrate out all nuisance parameters to recover the marginal
distribution we care about, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

@@

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, we rewrite the joint
probability distribution as a product of conditional distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}},
    \underline{\underline{R}}
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}, 
    \underline{\underline{R}}
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each of the terms. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. We invite the reader to check the supplementary material where we
work through each of these terms to build the posterior distribution to sample
from.

>>>>

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $T \times B$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique relevant barcodes we care
about (hereafter referred to as mutant barcodes), $M$, and the number of unique
neutral barcodes, $N$, and $T$ is the number of time points where measurements
were taken. The data matrix is then of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
$${#eq-R_mat}
where each row $\underline{r}_t$ is a $B$-dimensional array containing the
raw barcode counts at time $t$. We can further split each vector
$\underline{r}_t$ into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \;
\underline{r}_t^{M}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the neutral lineages barcode counts
$\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
counts $\underline{r}_t^{M}$. Following the same logic, matrix
$\underline{\underline{R}}$ can be split into two matrices as
$$
\underline{\underline{R}} = \left[ 
\underline{\underline{R}}^N \; \underline{\underline{R}}^M
\right],
$${#eq-R-mat_split}
where $\underline{\underline{R}}^N$ is a $T \times N$ matrix with the
barcode reads time series for each neutral lineage and
$\underline{\underline{R}}^M$ is the equivalent $T \times M$ matrix for the
mutant lineages.

Our objective is to compute the joint probability
distribution for all relative fitness values given our data. We can express this 
joint posterior distribution using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function, unless
otherwise stated. To develop our model, we can ignore the denominator on the
right-hand side of @eq-bayes_obj, since it serves as a normalization constant
that does not depend on our parameter of interest. However, computing this 
normalization constant represents a major challenge we will have to overcome. 
For now, we can simply write
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)
$${#eq-bayes_obj_propto}

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. We can always marginalize the joint
distribution to obtain the distribution we are interested in. To include these
parameters, let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R_mat, we can split $\underline{\underline{F}}$
into two matrices of the form
$$
\underline{\underline{F}} = \left[ 
\underline{\underline{F}}^N \; \underline{\underline{F}}^M
\right],
$${#eq-F-mat_split}
to separate the corresponding mutant and neutral barcode frequencies. With these
nuisance variables in hand, the full inference problem we must solve takes the
form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
We can numerically integrate out all nuisance parameters to recover the marginal
distribution we care about, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

@@

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, as depicted in @fig-01(C),
the statistical dependencies of our observations and latent variables allows us
to rewrite the joint probability distribution as a product of conditional
distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M,
    \underline{\underline{R}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N,
    \underline{\underline{R}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each of the terms. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. We invite the reader to check the supplementary material where we
work through each of these terms to build the posterior distribution to sample
from. We refer the reader to `[ref SI section]` for an extended description of
the model with specific functional forms for each term on the left-hand side of
@eq-split_posterior as well as `[ref SI on hierarchical]` where the model is
extended to account for multiple experimental replicates.

### Variational Inference

One of the technical challenges to the adoption of Bayesian methods is the
analytical intractability of integrals such as that of @eq-bayes_full_marginal.
Furthermore, even though efficient Markov Chain Monte Carlo (MCMC) algorithms
such as Hamiltonian Montecarlo can numerically perform this integration
[@betancourt2017], the dimensionality of the problem in @eq-split_posterior makes
an MCMC-based approach prohibitively slow.

To overcome this computational limitation, in this work, we rely on the recent
development of the automatic differentiation variational inference algorithm
(ADVI) [@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta \mid
\underline{\underline{R}})$, where $\phi$ fully parametrizes the approximate
distribution. As further explained in `[point to SI section]`, the numerical
integration problem is replaced by an optimization problem of the form
$$
q^*_\phi(\theta \mid \underline{\underline{R}}) = \min _\phi
D_{KL}(
    q_\phi(\theta \mid \underline{\underline{R}}) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the 
complicated high-dimensional numerical integration problem is transformed into
a gradient descent over the parameters of the approximate posterior distribution
$\phi$ such that it minimizes the KL divergence with respect to the true 
posterior distribution. Although to compute @eq-vi_objective we require the
posterior distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the 
variational free energy [@buckley2017]---is mathematically equivalent to 
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to `[point to SI section]` for a short primer on variational inference.

This work is accompanied by the Julia library `BayesFitness.jl` that makes use
of the implementation of both MCMC-based integration as well as ADVI
optimization to numerically approximate the solution of @eq-bayes_full_marginal
within the Julia ecosystem [@ge2018].

### 2023-09-06 doc/_experiment.qmd
making further progress on several sections

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time to then
dilute the culture back into fresh media and repeat for roughly 4-6 cycles,
depending on the initial abundances of the mutants, and their relative growth
rates. The DNA barcodes are sequenced at the end of each growth cycle to
quantify the relative abundance of each of the barcodes. We point the reader to
@kinsler2020 for specific details on these assays for *S. cerevisiae*.
@fig-01(B) presents a typical barcode trajectory where the darker barcode
trajectories represent the neutral lineages.

![**Typical competitive fitness experiment**. (A) Schematic of the typical
experimental design to determine the competitive fitness of an ensemble of
barcoded genotypes. Genotypes are pooled together and grown over multiple
growth-dilution cycles. At the end of each cycle, a sample is processed to
generate a library for amplicon sequencing. (B) Typical barcode trajectory
dataset. From each time point, the relative frequency of each barcode is
determined from the total number of reads. Different shades of blue represent
different relative fitness. Darker gray lines define the typical trajectory of
neutral lineages.](./figs/fig01){#fig-01}

>>>>

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time to then
dilute the culture back into fresh media and repeat for roughly 4-7 cycles,
depending on the initial abundances of the mutants, and their relative growth
rates. The DNA barcodes are sequenced at the end of each growth cycle to
quantify the relative abundance of each of the barcodes. We point the reader to
@kinsler2020 for specific details on these assays for *S. cerevisiae*.
@fig-01(B) presents a typical barcode trajectory where the darker barcode
trajectories represent the neutral lineages.

![**Typical competitive fitness experiment**. (A) Schematic of the typical
experimental design to determine the competitive fitness of an ensemble of
barcoded genotypes. Genotypes are pooled together and grown over multiple
growth-dilution cycles. At the end of each cycle, a sample is processed to
generate a library for amplicon sequencing. (B) Typical barcode trajectory
dataset. From each time point, the relative frequency of each barcode is
determined from the total number of reads. Different shades of blue represent
different relative fitness. Darker gray lines define the typical trajectory of
neutral lineages. (C) Bayesian network defining the statistical dependence
between observations (filled circles) and latent variables (empty circles). The
dotted line highlights the set of variables we are interested in. See
@sec-bayesian_inference for specific definitions of each
variable.](./figs/fig01){#fig-01}

### 2023-09-06 doc/_fitness_model.qmd
making further progress on several sections

## Preliminaries on mathematical notation
Before jumping directly into the Bayesian inference pipeline, let us establish
the mathematical notation to be used throughout this paper. We define
(column) vectors as underlined lowercase symbols such as
$$
\underline{x} = \begin{bmatrix}
    x_1\\
    x_2\\
    \vdots\\
    x_N
\end{bmatrix}.
$${#eq-vec_def}

In the same way, we define matrices as double-underline uppercase symbols such
as $\underline{\underline{A}}$.

@@

Empirically, the barcode relative frequency trajectories follow an exponential
function of the form
$$
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
the relative fitness with respect to the reference strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. We can rewrite
@eq-fitness as
$$
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
$${#eq-logfreq}
@eq-logfreq separates the observables---the barcode frequencies---from the
latent parameters we want to infer from the data---the population mean fitness
and the barcode relative fitness. This is ultimately the functional form used in
our inference pipeline. Therefore, the relative fitness is computed by knowing
the barcode log frequency ratio of each barcode throughout the growth-dilution
cycles.

@@

Our ultimate objective is to infer the relative fitness $s^{(m)}$ for each of
the $M$ mutants in the experiment. To do so, we account the three main sources
of uncertainty in our model:

1. Uncertainty in the determination of frequencies. Our model relates
frequencies between adjacent time points to the fitness of the corresponding
strain. However, we do not directly measure frequencies. Instead, our data for
each barcode consists of a vector of counts $\underline{r}^{(b)}_t$ for each of
the $T$ time points in which the measurements were taken.
2. Uncertainty in the value of the population mean fitness. We define neutral
lineages to have fitness $s^{(n)} = 0$, helping us anchor the value of the
population mean fitness $\bar{s}_t$ for each pair of adjacent time points. 
Moreover, we take this parameter as an empirical observable from the data, 
meaning that we do not have a functional form that relates $\bar{s}_t$ to
$\bar{s}_{t+1}$. Thus, we must infer the $T-1$ values of this population mean 
fitness with their uncertainty that must be propagated to the value of
the mutants' relative fitness.
3. Uncertainty in each of the mutants' fitness values.

To account for all these sources of uncertainty in a principled way, we develop
a Bayesian inference pipeline.

>>>>

## Preliminaries on mathematical notation
Before jumping directly into the Bayesian inference pipeline, let us establish
the mathematical notation used throughout this paper. We define
(column) vectors as underlined lowercase symbols such as
$$
\underline{x} = \begin{bmatrix}
    x_1\\
    x_2\\
    \vdots\\
    x_N
\end{bmatrix}.
$${#eq-vec_def}
In the same way, we define matrices as double-underline uppercase symbols such
as
$$
\underline{\underline{A}} =
\begin{bmatrix}
    A_{11} & A_{12} & \cdots & A_{1N}\\
    A_{21} & A_{22} & \cdots & A_{2N}\\
    \vdots & \vdots & \ddots & \vdots\\
    A_{M1} & A_{M2} & \cdots & A_{MN}\\
\end{bmatrix}.
$${#eq-mat_def}

@@

Empirically, the barcode relative frequency trajectories follow an exponential
function of the form
$$
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
the relative fitness with respect to the reference strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. We can rewrite
@eq-fitness as
$$
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
$${#eq-logfreq}
@eq-logfreq separates the observables---the barcode frequencies---from the
latent parameters we want to infer from the data---the population mean fitness
and the barcode relative fitness. This is ultimately the functional form used in
our inference pipeline. Therefore, the relative fitness is computed by knowing
the log frequency ratio of each barcode throughout the growth-dilution cycles.

@@

Our ultimate objective is to infer the relative fitness $s^{(m)}$ for each of
the $M$ relevant barcodes in the experiment. To do so, we account for the three
primary sources of uncertainty in our model:

1. Uncertainty in the determination of frequencies. Our model relates
frequencies between adjacent time points to the fitness of the corresponding
strain. However, we do not directly measure frequencies. Instead, our data for
each barcode consists of a length $T$ vector of counts $\underline{r}^{(b)}$ for
each of the $T$ time points in which the measurements were taken.
2. Uncertainty in the value of the population mean fitness. We define neutral
lineages to have fitness $s^{(n)} = 0$, helping us anchor the value of the
population mean fitness $\bar{s}_t$ for each pair of adjacent time points. 
Moreover, we take this parameter as an empirical observable from the data, 
meaning that we do not have a functional form that relates $\bar{s}_t$ to
$\bar{s}_{t+1}$. Thus, we must infer the $T-1$ values of this population mean 
fitness with their uncertainty that must be propagated to the value of
the mutants' relative fitness.
3. Uncertainty in each of the mutants' fitness values.

To account for all these sources of uncertainty in a principled way, in the next
section, we develop a Bayesian inference pipeline.

### 2023-09-06 doc/_intro.qmd
making further progress on several sections

The advent of DNA barcoding and high-throughput sequencing has opened new venues
to understand microbial evolutionary dynamics with an unprecedented level of
temporal resolution [@levy2015; @nguyenba2019a; @ascensao2023]. These
experimental efforts rely on our ability to reliably infer the relative fitness
of an ensemble of diverse genotypes. Moreover, inferring these fitness values
over an ensemble of environmental conditions can help us determine the
phenotypic diversity of a rapid adaptation process @kinsler2020.

As with any other sequencing-based quantification, tracking lineages via DNA
barcode sequencing is inexorably accompanied by noise sources coming from
experimental manipulation of the microbial cultures, DNA extraction and
sequencing library preparation that involves multiple rounds of PCR
amplification, and the sequencing process itself. Thus, accounting for the
uncertainty when inferring the parameters we care about from the data is a key
step to drawing reliable conclusions. Bayesian statistics presents a paradigm
with which to account for all known sources of uncertainty in a principled way
@eddy2004a. This, combined with the development of modern Markov Chain Monte
Carlo sampling algorithms have boosted a resurgence of Bayesian methods in
different fields @efron2013a.

We present a Bayesian inference pipeline to properly quantify our uncertainty
about the parametric information we can extract from high-throughput competitive
fitness assays. In these assays, the fitness of an ensemble of genotypes is
determined relative to a reference genotype [@kinsler2020; @ascensao2023] [`add
reference to fig`]. Small changes to our model allow us to account for relevant
experimental variations of these competition assays: determination of multiple
fitness values per genotype for growth-dilution cycles on different
environments, and a hierarchical model to account for information from
independent experimental replicates. We benchmark the ability of our pipeline to
infer fitness parameters against synthetic data with known ground truth and
re-analyze published data from @kinsler2020 to assess the method's performance
on real experimental measurements. The method is accompanied with a `Julia`
package to readily implement the inference pipeline with state-of-the-art
scientific computing software. [`here mention something about the Python wrapper
if appropriate`].

>>>>

The advent of DNA barcoding and high-throughput sequencing has opened new venues
to understand microbial evolutionary dynamics with an unprecedented level of
temporal resolution [@levy2015; @nguyenba2019a; @ascensao2023]. These
experimental efforts rely on our ability to reliably infer the relative fitness
of an ensemble of diverse genotypes. Moreover, inferring these fitness values
over an ensemble of environmental conditions can help us determine the
phenotypic diversity of a rapid adaptation process [@kinsler2020].

As with any other sequencing-based quantification, tracking lineages via DNA
barcode sequencing is inexorably accompanied by noise sources coming from
experimental manipulation of the microbial cultures, DNA extraction and
sequencing library preparation that involves multiple rounds of PCR
amplification, and the sequencing process itself. Thus, accounting for the
uncertainty when inferring the parameters we care about from the data is a key
step to drawing reliable conclusions. Bayesian statistics presents a paradigm
with which to account for all known sources of uncertainty in a principled way
[@eddy2004a]. This, combined with the development of modern Markov Chain Monte
Carlo sampling algorithms [@betancourt2017] and approximate variational
approaches [@kucukelbir2016] have boosted a resurgence of Bayesian methods in
different fields [@efron2013a].

We present a Bayesian inference pipeline to properly quantify our uncertainty
about the parametric information we can extract from high-throughput competitive
fitness assays. In these assays, the fitness of an ensemble of genotypes is
determined relative to a reference genotype [@kinsler2020; @ascensao2023] (See
@fig-01 for a schematic depiction of the experimental setup). Small changes to
our model allows us to account for relevant experimental variations of these
competition assays: determination of multiple fitness values per genotype for
growth-dilution cycles on different environments, a hierarchical model to
account for information from independent experimental replicates, and a
hierarchical model for a within-experiment structure where multiple independent
barcodes map to a specific genotype. We benchmark the ability of our pipeline to
infer fitness parameters against synthetic data with known ground truth and
re-analyze published data from @kinsler2020 to assess the method's performance
on real experimental measurements. The method is accompanied by a `Julia`
package to readily implement the inference pipeline with state-of-the-art
scientific computing software. [`here mention something about the Python wrapper
if appropriate`].

### 2023-09-06 doc/_single_results.qmd
making further progress on several sections

To assess the inference pipeline performance we applied it to a simulated
dataset with known ground truth (See `[ref to SI]` for details on simulation).
@fig-02(A) shows the structure of the synthetic dataset. The majority of
barcodes of interest (dim color lines) are adaptive compared to the neutral
barcodes ($s^{(m)} > 0$). This experimental condition is required to reliably
measure the relative fitness since all barcodes of interest occupy a small
fraction of the population at the beginning of the experiment `[ref SI section
on detrimental mutants]`. Although the barcode frequency trajectories look
relatively smooth, our fitness model requires the computation of the log
frequency ratio between adjacent time points. @fig-02(B) shows such data
transformation. Here, we can better appreciate the variability in the input that
goes into our model. This is especially clear for the darker lines representing
the neutral barcodes since they are supposed to be identically distributed under
the log frequency ratio transformation.

A way to visualize the performance of our inference pipeline in fitting our
fitness model to the observed data is to compute the so-called posterior
predictive checks (PPC). In short, the PPC consists of bootstrapping the
generation of synthetic datasets in agreement with the results from the
inference results. In other words, we use the resulting parameter values from
the MCMC chain to generate possible datasets in agreement with the inferred
values. @fig-02(C) shows these results for all neutral lineages (upper left
corner plot) and a few representative non–neutral barcodes. The dark color line
shows the median value we would expect the data to be found, while the shaded
regions show the 68\% and 95\% credible region, i.e. the regions where we expect
to find the data with the corresponding probability---or in terms of our
parameter, the $X\%$ credible region is the interval where we expect the true
parameter to lie with $X\%$ probability. The main advantage of our method is the
natural interpretability of these credible regions. A common mistake in the
literature is to confuse frequentist confidence intervals with Bayesian credible
regions when they are not equivalent @morey2016.

To capture the global performance of the model, @fig-02(D) compares the known
groud truth with the inferred relative fitness value for all barcodes of
interest. There is a great degree of correspondence between these values, with
the error bars, representing the 68\% credible region for the parameter value
crossing the identity line for most barcodes. This latter point is made clear
with @fig-02(E) where the empirical cumulative distribution function (ECDF) of
the absolute differences between the median inferred value and the true value
shows that all median inferred values are within 0.3 units of relative fitness
from the ground truth. Furthermore, because of the Bayesian scope of our method,
we do not need to limit ourselves to a point estimate of this quantity, but
report (asymmetric) error bars that fully quantify our degree of belief in the
inferred value as shown on top of the panels in @fig-02(C).

>>>>

To assess the inference pipeline performance we applied it to a simulated
dataset with known ground truth relative fitness values (See `[ref to SI]` for
details on simulation). @fig-02(A) shows the structure of the synthetic dataset.
The majority of barcodes of interest (dim color lines) are adaptive compared to
the neutral barcodes ($s^{(m)} > 0$). This experimental condition is required to
reliably measure the relative fitness since all barcodes of interest occupy a
small fraction of the population at the beginning of the experiment `[ref SI
section on detrimental mutants]`. Although the barcode frequency trajectories
look relatively smooth, our fitness model requires the computation of the log
frequency ratio between adjacent time points as derived in @eq-logfreq.
@fig-02(B) shows such data transformation where we can better appreciate the
observational noise input into our statistical model. This noise is
evident for the darker lines representing the neutral barcodes since
all of these lineages are assumed to be identically distributed under this log
frequency ratio transformation.

A way to visualize the performance of our inference pipeline in fitting our
fitness model to the observed data is to compute the so-called posterior
predictive checks (PPC). In short, the PPC consists of bootstrapping the
generation of synthetic datasets in agreement with the results from the
inference results. In other words, we use the resulting parameter values from
the ADVI inference to generate possible datasets in agreement with the inferred
values. @fig-02(C) shows these results for all neutral lineages (upper left
corner plot) and a few representative non-neutral barcodes. The dark color line
shows the median value we would expect the data to be found, while the shaded
regions show the 68\% and 95\% credible regions, i.e., the regions where we
expect to find the data with the corresponding probability---or in terms of our
parameter, the $X\%$ credible region is the interval where we expect the true
parameter to lie with $X\%$ probability. The main advantage of our method is
this natural interpretability of these credible regions since a common mistake
in the literature is to interpret frequentist confidence intervals as Bayesian
credible regions when they are not equivalent [@morey2016].

To capture the global performance of the model, @fig-02(D) compares the known
ground truth with the inferred relative fitness value for all barcodes of
interest. There is an excellent degree of correspondence between these values,
with the error bars, representing the 68\% credible region for the parameter
value crossing the identity line for most barcodes. This latter point is made
clear with @fig-02(E) where the empirical cumulative distribution function
(ECDF) of the absolute differences between the median inferred value and the
actual value shows that all median inferred values are within 0.3 units of
relative fitness from the ground truth.
